---
title: "Object Oriented Programming With R"
subtitle: "A walkthrough from Procedural to OOP"
author: "Adam Slack"
date: "2017-6-16"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## What is Object Oriented Programming?

- Object Oriented Programming (OOP) is a programming paradigm

- Involves the grouping of data with functions
    + These functions are used to apply changes to the contained data

- A collection of Data and Functions is called a class

- A Class is a blue print, from which instances of an object can be made

## Pros and Cons of OOP

Like all programming methodologies, there are upsides and downsides

Pros                                | Cons
------------------------------------|--------------------------
Scalable and Manageable Code bases  | Can be time consuming to design and produce
Less Code Redundancy                | Not always worth the effor for small projects
Portable and Re-usable code         |
Protects data from unwanted changes |

## Functionality vs Data
* Think about a program that involves cars
    + What can a car do? (The functionality)
    + What information you can record about a car? (The data)
  
Functionality       | Recordable Information
--------------------| ----------------------
Refuel              | Fuel
Drive               | Max Speed
Check fuel level    | Current Speed
check current speed | Colour

## Where does this fit in with code?

The information and functionality of a car in your program is specified inside of a class

A Class is composed of two sections

* Methods: 
    + What can the object do?
* Attributes:
    + what information is stored with the object?
    
Each Class can be used to instantiate objects

## A Class vs Objects

- If a class is a blueprint, an object is a physical instance of it

- A class can be used to create as many instances as you like
    - Same functionality
    - Same Data members
    - Different values for the data.

- Think of 3 different cars
    + Fiat Punto, Volkswagen Golf', Honda Jazz

- Each instance is a car
    + Same functionality
    + Same information can be recorded
    + At anyone time each car will have different values
    
## Standard R procedural Code
```{r echo = FALSE, results='hide',warning=FALSE,message=FALSE,error=FALSE}
# 
# R demonstration of Standard Procedural code.
# 

invisible(library(ggplot2))
invisible(library(reshape2))
library(R6)

```

```{r echo = TRUE}
cars <- c('Punto', 'Golf', 'Jazz')
colour <- c('beige', 'gold', 'neon pink')
fuel <- c(100,100,100) # inital percent of the fuel tank that is full
ppm <- c(10,20,5) # percent per mile fuel usage
```

```{r echo = TRUE}
remaining <- c() # what percentage of the fuel tank remains
miles <- 3 # how many miles do you plan to travel
```

```{r echo = TRUE}
for (i in 1:length(fuel)) { # calculate the new fuel tank values
  remaining <- c(remaining, (fuel[i] - (ppm[i]*3)))
}
```

```{r echo = TRUE}
car_frame <- data.frame(cars, fuel, remaining) # create a data frame
car_frame.m <- melt(car_frame, id.vars='cars') # melt the data frame for plotting
```

```{r echo = TRUE}
carplot <- ggplot(car_frame.m, aes(cars, value)) +   
  geom_bar(aes(fill = variable), position = "dodge", stat="identity")

```
## The resulting plot
```{r echo = FALSE, fig.keep=carplot}
carplot
```

## Adding Some functions
- The previous code can be improved by adding some functions.

- That way we can repeat what we did by calling the functions.

- We don't need to copy, paste and change the code that way.

## Drive Car
1. For the amount of time specified in distance
2. Calculate the change in fuel level
    - If the fuel goes below 0% dont record the change
3. Return the changes in fuel usage

```{r echo = TRUE}
drive_car <- function(starting_tank, ppm, distance) {
  tank_changes = c() # track the fuel usage using what % remains
  for(i in 0:distance) {
    new_tank <- starting_tank - (ppm * i)
    if(new_tank >= 0) {
      tank_changes <- c(tank_changes, new_tank) # only spend fuel if there was fuel to spend.
    }
  }
  return(tank_changes) # return the fuel usage
}

```

## Refuel Car
1. Record the amount of fuel added 
2. Cap the fuel level at 100%
3. Return the vector of fuel level changes

```{r echo = TRUE}
# creating a function to refuel a car
refuel_car <- function(tank_changes, tank_percentage) {
  new_tank <- tank_changes[length(tank_changes)] + tank_percentage
  if(new_tank > 100) {
    new_tank <- 100 # cant have more than a 100% full tank.
  }
  return(c(tank_changes, new_tank))
}

```
## Some data oriented functions
- Some functions will handle more abstract (or data oriented) aspects

- Padding of vectors in a collection such that they have equal length

- the construction of a dataframe so that plotting can be performed

## Pad Lengths
1. get the length of the longest vector
2. set the lengths of each vector to be that length
```{r echo=TRUE}
pad_lengths <- function(...) {
  arguments <- list(...)
  max_length <- max(lengths(arguments))
  args <- lapply(arguments, function(x, max_len) {
    length(x) <- max_length
    return(x)
  }, max_len = max_length)
  return(args)
}
```
## Create Data Frame
1. Pad vectors passed to the function
2. Create the data frame and set the column names
3. Transform and 'melt' the dataframe
```{r echo=TRUE}
create_data_frame <-function(car_names, ...) {
  car_data <- list(...)

  padded_fuel <- do.call(pad_lengths, car_data)
  car_frame <- data.frame(padded_fuel)
  colnames(car_frame) <- car_names
  
  car_frame.m <- melt(t(car_frame), id.vars="Cars", 
                                    value.name ="Fuel",
                                    variable.name = "Time")
  colnames(car_frame.m) <- c('Cars', 'Time', 'Fuel')
  return(car_frame.m)
}
```

## Using the functions
- You now have some functions
- These can be used, shared and adapted
    - Reduced risk of side effects
```{r echo=TRUE}
jazz_fuel_hist <- drive_car(100, 7.5, 10)
```

```{r echo = TRUE}
golf_fuel_hist <- drive_car(100, 2, 10)
golf_fuel_hist <- c(golf_fuel_hist,
                    drive_car(golf_fuel_hist[length(golf_fuel_hist)],
                              5, 10))
```

```{r echo = TRUE}
punto_fuel_hist <- drive_car(100, 10, 20)
punto_fuel_hist <- refuel_car(punto_fuel_hist, 70)
punto_fuel_hist <- c(punto_fuel_hist, drive_car(70, 10, 20))

```

## Using the functions
```{r echo=TRUE}
car_data_frame <- create_data_frame(car_names = c('Punto',
                                                  'Golf',
                                                  'Jazz'),
                                   punto_fuel_hist,
                                   golf_fuel_hist,
                                   jazz_fuel_hist)
```

```{r echo = TRUE}
fuel_line_plot <-ggplot(car_data_frame, 
                        aes(x = Time,
                            y = Fuel,
                            group  = Cars,
                            colour = Cars)) +   
                        geom_line() + 
                        geom_point()
```

## The resulting plot
```{r echo = FALSE, fig.keep=carplot, warning=FALSE,message=FALSE,error=FALSE}
fuel_line_plot
```


## This Isn't Perfect

- There is always way to improve Code
- The Drive function could be adapted
    - allow the user to pass a function that determines fuel usage
    - means it doesn't have to be a fixed linear usage
- The data is still being acted upon by functions
    - they could be grouped and protected further


## R6 Classes
- R has a range of differnt solutions to object oriented programming
- R6 provides a similar style to many other languages.
    - Private Data Members and Methods
    - Public Data Members and Methods

## R6 Classes
- An R6 class is essentially a pair of lists
- One list contains elements easily accessible
- The other list contains elements protected from direct access
- These are referred to as 'public' and 'private'

## R6 Classes
- In each list Data members and methods are defined
- A constructor can be defined in the public list
    - This is a special method that executes straight away
    - Invoked whenever an instance is made.
- The constructor allows you to initialise data members

## Template Class
```{r echo=TRUE}
ClassName <- R6Class('ClassName',
                  public = list(
                  public_data = NULL,
                    
                  initialize = function(public_data, private_data) {
                    self$public_data <- public_data
                    private$private_data <- private_data
                  },
                  
                  public_method = function() {}
                ),
                
                private = list(
                  private_data = NULL,
                  private_method = function() { }
                )
)
```

## Car implementation
- When implementing a Class you should think about a few aspects
    - What functionality should my instances have?
    - What Information should i know about each instance?
    - What Information should be hidden from me?
    
## Car Implementation
Data Members

-Public
    - Name
    - Current Amount of Fuel (Gallons)
    - History of Fuel Usage (Gallons/minute)
    - Current Speed (Miles/hour)
-Private
    - Max Speed (Miles/hour)
    - Max Amount of Fuel (Gallons)
    
## Car Implementaion
Methods

- Public
    - Class Constructor (Initialize)
    - Get Max Speed
    - Get Max Fuel
    - Drive
    - Refuel
    
- Private
    - None...

## Public Data Members 
- Start off defining Public Data members and methods
    - Data Members defined first, followed by the methods
```{r echo=TRUE, eval=FALSE}
Car <- R6Class('Car',
               public = list(
                 name = NULL,
                 fuel_usage = NULL,
                 current_speed = 0,
                 current_tank = 0,
```

## Class Constructor
- The constructor should be the first method defined
- This is called whenever a new instance of the class is made.
```{r echo=TRUE, eval=FALSE}

   ## CLASS CONSTRUCTOR
   initialize = function(name, max_speed, max_tank, current_tank) {
     self$name          <- name
     self$fuel_usage    <- c(current_tank)
     self$current_tank  <- current_tank
     private$max_speed  <- max_speed
     private$max_tank   <- max_tank
   },
```

## Public Getters
- Used to get the value of a private data member
- Means you can See the value, with no risk of altering it.
```{r echo=TRUE, eval=FALSE}
   get_max_speed = function() {
     return(private$max_speed )
   },
```

```{r echo=TRUE, eval=FALSE}

   get_max_tank = function() {
     return(private$max_speed)
   },
```

## Public Methods - Drive
```{r echo=TRUE, eval=FALSE}
   drive = function(mph, distance, mpg) {
     if(mph > private$max_speed) {
       mph <- private$max_speed
     }
     mpm <- mph/60
     journey_time <- distance/mpm
     gallons_used <- distance/mpg
     gallon_pm <- gallons_used/journey_time

     for(i in 1:journey_time) {
       self$current_tank = self$current_tank - gallon_pm
       
       if(self$current_tank < 0) {
         self$current_tank = 0 .
       }
       self$fuel_usage <- c(self$fuel_usage, self$current_tank)  
     }
     return(self$fuel_usage)
   },
```

## Public Methods - Refuel
```{r echo=TRUE, eval=FALSE}
   refuel = function(gallons_added) {
     self$current_tank <- self$current_tank + gallons_added
     if(self$current_tank > private$max_tank) {
       self$current_tank <- private$max_tank
     }
     self$fuel_usage <- c(self$fuel_usage, self$current_tank)
   }
 ),

```

## Private Data
- Private data Members and methods should be defined after public
- Private Data Members should be defined before methods though.
```{r echo=TRUE, eval=FALSE}

   private = list(
     # PRIVATE DATA MEMBERS
     max_speed=NULL,
     max_tank=NULL
   ))
```

```{r echo=FALSE}
Car <- R6Class('Car',
               public = list(
                 # PUBLIC DATA MEMBERS
                 name = NULL,
                 fuel_usage = NULL,
                 current_speed = 0,
                 current_tank = 0,

                 ## CLASS CONSTRUCTOR
                 initialize = function(name, max_speed, max_tank, current_tank) {
                   self$name          <- name
                   self$fuel_usage    <- c(current_tank)
                   self$current_tank  <- current_tank
                   private$max_speed  <- max_speed
                   private$max_tank   <- max_tank
                 },
                 
                 ## GETTERS
                 get_max_speed = function() {
                   return(private$max_speed )
                 },
                 
                 get_max_tank = function() {
                   return(private$max_speed)
                 },
                 
                 ## PUBLIC METHODS
                 drive = function(mph, distance, mpg) {
                   if(mph > private$max_speed) {
                     mph <- private$max_speed
                   }
                   mpm <- mph/60
                   journey_time <- distance/mpm
                   gallons_used <- distance/mpg
                   gallon_pm <- gallons_used/journey_time

                   for(i in 1:journey_time) {
                     self$current_tank = self$current_tank - gallon_pm
                     
                     if(self$current_tank < 0) {
                       self$current_tank = 0 # only spend fuel if there was fuel to spend.
                     }
                     self$fuel_usage <- c(self$fuel_usage, self$current_tank)  
                   }
                   return(self$fuel_usage)
                 },
                 
                 refuel = function(gallons_added) {
                   self$current_tank <- self$current_tank + gallons_added
                   if(self$current_tank > private$max_tank) {
                     self$current_tank <- private$max_tank
                   }
                   self$fuel_usage <- c(self$fuel_usage, self$current_tank)
                 }
               ),
               private = list(
                 # PRIVATE DATA MEMBERS
                 max_speed=NULL,
                 max_tank=NULL
               ))


DrivePlotProducer <- R6Class('PlotProducer',
                             
                             public = list(
                               car_names = c(),
                               
                               pad_lengths = function(...) {
                                 arguments <- list(...)
                                 max_length <- max(lengths(arguments))
                                 
                                 args <- lapply(arguments, function(x, max_len) {
                                   length(x) <- max_length
                                   return(x)
                                 }, max_len = max_length)
                                 return(args)
                               },
                               
                               create_data_frame = function(...) {
                                 car_data <- list(...)
                                 
                                 padded_fuel <- do.call(self$pad_lengths, car_data)
                                 car_frame <- data.frame(padded_fuel)
                                 colnames(car_frame) <- self$car_names
                                 
                                 car_frame.m <- melt(t(car_frame), id.vars="Cars", 
                                                     value.name ="Fuel",
                                                     variable.name = "Time")
                                 colnames(car_frame.m) <- c('Cars', 'Time', 'Fuel')
                                 return(car_frame.m)
                               },
                               
                               create_plot = function(...) {
                                 car_objects <- list(...)
                                 fuel_usages <- lapply(car_objects, function(x){return(x$fuel_usage)})
                                 self$car_names <- lapply(car_objects, function(x){return(x$name)})
                                 car_frame <- do.call(self$create_data_frame, fuel_usages)
                                 
                                return(ggplot(car_frame, aes(x = Time,
                                                                             y = Fuel,
                                                                             group  = Cars,
                                                                             colour = Cars)) +   
                                   geom_line() + geom_point())
                               }
                             )
                          )

```

## The Class In Use
Creating Instances of a Class to use.

- Call the 'new' function associated with a class
- Pass values as parameters to the class
    - These are used in the initialisation/construction of the object
```{r echo=TRUE}
golf <- Car$new('Golf', 100, 10, 3)

```
- We now have a golf object
    - All the functionality and data associated with a car.
```{r echo=TRUE, results='hide',warning=FALSE,message=FALSE,error=FALSE}
golf$drive(10,10,10)
golf$drive(mph=40, distance=10, mpg=30)
```

## Inspecting the Instance

```{r echo=TRUE}
golf
```

## Creating More Instances
```{r echo=TRUE, results='hide',warning=FALSE,message=FALSE,error=FALSE}
golf <- Car$new('Golf', 100, 10, 3)
golf$drive(10,10,10)
golf$drive(mph=40, distance=10, mpg=30)
```

```{r echo=TRUE, results='hide',warning=FALSE,message=FALSE,error=FALSE}
jazz <- Car$new('Jazz', 200, 10, 10)
jazz$drive(30, 2, 40)
jazz$drive(40, 1, 38)
jazz$drive(70, 10, 44)
jazz$drive(110, 60, 30)
```

```{r echo=TRUE, results='hide',warning=FALSE,message=FALSE,error=FALSE}
fiat <- Car$new('Fiat', 90, 8, 1) # They have a fuel leak...
fiat$drive(30, 3, 1)
fiat$refuel(5) # Refueled to get to the garage
fiat$drive(30, 0.5, 1)
fiat$drive(40, 5, 40) # Fuel Leak Repaired
fiat$drive(30, 2, 45)
fiat$drive(60, 10, 45)
```

## What about the plotting aspect?
- These Cars are quite straight forward to picture and visualise

More abstract classes can exist.

- Objects that have behaviour relating to:
    - Plotting Figures
    - Analyising Data
    - Producing Datasets
    
## Plotting Car Fuel Data
An Object specifically designed to plot Car Fuel Data.

- Needs
    - **Data Member:** Car Names
    - **Method:** Plot Data
    - **Method:** Pad Vectors
    - **Method:** Create Data Frame

- Could also Have
    - Data members for specifying plot details...
    - methods for handling specific plot details...
    
## Implementing a Car Plotter Object

- Start Defining public Data Members

```{r echo=TRUE, eval=FALSE}
DrivePlotProducer <- R6Class('PlotProducer',
                             
                             public = list(
                               car_names = c(),
```

## Pad Vectors
- Then begin implementing public Methods
```{r echo=TRUE, eval=FALSE}                             
   pad_lengths = function(...) {
     arguments <- list(...)
     max_length <- max(lengths(arguments))
     
     args <- lapply(arguments, function(x, max_len) {
       length(x) <- max_length
       return(x)
     }, max_len = max_length)
     return(args)
   },

```

## Create Data Frame
```{r echo=TRUE, eval=FALSE}                             
create_data_frame = function(...) {
   car_data <- list(...)
   
   padded_fuel <- do.call(self$pad_lengths, car_data)
   car_frame <- data.frame(padded_fuel)
   colnames(car_frame) <- self$car_names
   
   car_frame.m <- melt(t(car_frame), id.vars="Cars", 
                       value.name ="Fuel",
                       variable.name = "Time")
   colnames(car_frame.m) <- c('Cars', 'Time', 'Fuel')
   return(car_frame.m)
  },
```

## Create Plot
```{r echo=TRUE, eval=FALSE}                             
 create_plot = function(...) {
       car_objects <- list(...)
       fuel_usages <- lapply(car_objects, function(x){return(x$fuel_usage)})
       self$car_names <- lapply(car_objects, function(x){return(x$name)})
       car_frame <- do.call(self$create_data_frame, fuel_usages)
       
      return(ggplot(car_frame, aes(x = Time,
                                                   y = Fuel,
                                                   group  = Cars,
                                                   colour = Cars)) +   
         geom_line() + geom_point())
     }
   )
)

```

## Using Create Plot
- Create an Instance
```{r echo=TRUE, results='hide',warning=FALSE,message=FALSE,error=FALSE}
plot_producer <- DrivePlotProducer$new()

```

- Then use the functionality of it
```{r echo=TRUE, results='hide',warning=FALSE,message=FALSE,error=FALSE}

car_plot <- plot_producer$create_plot(golf, jazz, fiat)

```

## The Output
```{r echo = FALSE, fig.keep=carplot, warning=FALSE,message=FALSE,error=FALSE}
car_plot
```
